Computer-Assisted Analysis of the \\ Anderson-Hájek Ontological Controversy

### 2014-10-26 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
fixed some typos

\noindent The axioms in G\"odel's ontological proof
\citep{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) entail
what is called \emph{modal collapse}
\citep{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended. This fact, which
has recently been confirmed with higher-order automated theorem
provers \citep{C40,J30}, has led to strong criticism of the argument
and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took
greater care of the semantics of higher-order quantifiers
in the presence of modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof}, \citet{Hajek2002}
and \citet{bjordal99}, proposed emendations of G\"odel's
axioms and definitions. They require neither comprehension
restrictions nor more complex semantics. Therefore, they are
technically simpler to analyze with computer support. We have
formalized them using the proof assistant Isabelle/HOL \citep{Isabelle}
together with the automated higher-order reasoners Leo-II \citep{C26},
Satallax \citep{brown2012satallax}, Metis 
\citep{Hurd03first-orderproof}, and Nitpick \citep{Nitpick}.   Our
formalizations\footnote{The   formalizations are available in the
subdirectories \url{Anderson}, \url{Hájek} and \url{Bjordal} at \url{gi
thub.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/
}.} employ the embedding of higher-order modal logic (HOML) in
classical higher-order logic (HOL) as introduced in previous work
\citep{C40,J30,J23}. We explored the effect of different domain
conditions on the provability of lemmas, theorems and even axioms.
This was motivated by a controversy between Hájek and Anderson
regarding the redundancy of some axioms in Anderson's emendation. In
\emph{constant domain semantics}, the individual domains are the same
in all possible worlds. In \emph{varying domain semantics}, the
domains may vary from world to world. This variation is technically
encoded with the help of an existence relation expressing which
individuals actually exist in each world. Quantifiers are then
uniformly formalized as \emph{actualistic quantifiers} (i.e. guarded by
the existence relation). Our main results are summarized here.

@@

The redundancy of A4 and A5 is particularly controversial. Magari \cite{Magari1988} claimed that A4 and A5 are superfluous\footnote{
  An axiom $A$ is superfluous w.r.t. a set of axioms $S$ iff T3 is entailed by $S \setminus \{ A \}$.
}, arguing that T3 is true in all models of the other axioms and definitions by Gödel. \citet[p.~5-6]{Hajek_Magari_and_others_1996} investigated this further, refuted Magari's claim, but claimed that A4 and A5 are indeed superfluous under the assumption of an additional axiom (PEP, pointwise equality for positiveness), which Magari probably assumed silently. Moreover, Hájek \citet[p.~2]{Hajek_Magari_and_others_1996} cites his earlier work\footnote{Although \citep{Hajek_der_Mathematiker_2001} precedes \citep{Hajek_Magari_and_others_1996} in writing, it was published only 5 years later, in German.} \citep{Hajek_der_Mathematiker_2001}, where he claims (in Theorem 5.3) that for Anderson's emended theory \citep{anderson90:_some_emend_of_goedel_ontol_proof}, A4 and A5 are not only superfluous, but also redundant. \citet[footnote 1 in p.~1]{AndersonGettings}, in a footnote, rebutted Hájek's claim, arguing that the redundancy of A4 and A5 holds only under constant domain semantics, while Anderson's emended theory ought to be taken under Cocchiarella's semantics \citep{Cocchiarella} (a varying domain semantics). Our results show that Hájek was originally right, under both constant and varying domain semantics.

Nevertheless, \citet[p.~7]{Hajek2002} (in page 7) acknowledges Anderson's rebuttal, and apparently accepts it, as evidenced by  his use\footnote{
  A4 and A5 are used by \citet[p.~11]{Hajek2002} (page 11) in the proofs of, respectively, Lemma 4 and Theorem 4.
} 
of A4 and A5, as well as varying domain semantics, in his new emendation (named \AOEH \citep[sec.~4]{Hajek2002}, cf. Appendix \ref{apx:Hajek1}), which replaces Anderson's A:A1 and A2 by a simpler axiom H:A12. Surprisingly, the computer-assisted formalization of \AOEH shows that A4 and A5 are superfluous (not needed for proving T3). Moreover, A4 and A5 are independent\footnote{
  An axiom $A$ is \emph{independent} of a set of axioms $S$ iff there are models of $S$ where $A$ is true and other models of $S$ where $A$ if false. 
} of the other axioms and definitions. Therefore, A4 and A5 are not redundant, despite their superfluousness.

Although, Hájek did not notice the superfluousness of A4 and A5 in his \AOEH, he did describe yet another emendation (his \AOEHH, cf. Appendix \ref{apx:Hajek2}) where A4 and A5 are superfluous (though no claim is made w.r.t. to their redundancy), if A3 is replaced by a stronger axiom (H:A3) additionally stating that the property of actual existence is positive \cite[sec.~5]{Hajek2002}. Formalization of \AOEHH shows that A4 is not only superflous, but also redundant. For A5, no conclusive results were achieved; neither a proof nor a countermodel could be automatically generated. Surprisingly, a countermodel for the weaker A3 was succesfully generated. This is somewhat unsatisfactory (for theistic goals), because it shows that \AOEHH does not entail the positiveness of being God-like.

Nevertheless, \AOEHH is explicitly regarded by \citet[p.~12]{Hajek2002} (page 12) as just an intermediary step towards a more natural theory, based on a more sophisticated notion of positiveness. That is his final emendation (\AOEHHH, cf. \ref{apx:Hajek3}), which restores A3 and does use A4 and A5, albeit in a modified form (i.e. H:A4 and H:A5). The formalization of \AOEHHH shows that H:A4 is independent. The old A5 is independent as well, and both H:A4 and H:A5 are superfluous, but no conclusive results were achieved regarding independence or redundancy of H:A5.

@@

\printbibliography

>>>>

\noindent The axioms in G\"odel's ontological proof
\citep{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) entail
what is called \emph{modal collapse}
\citep{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended. This fact, which
has recently been confirmed with higher-order automated theorem
provers \citep{C40,J30}, has led to strong criticism of the argument
and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took
greater care of the semantics of higher-order quantifiers
in the presence of modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof}, \citet{Hajek2002}
and \citet{bjordal99}, proposed emendations of G\"odel's
axioms and definitions. They require neither comprehension
restrictions nor more complex semantics. Therefore, they are
technically simpler to analyze with computer support. We have
formalized them using the proof assistant Isabelle/HOL \citep{Isabelle}
together with the automated higher-order reasoners Leo-II \citep{C26},
Satallax \citep{brown2012satallax}, Metis 
\citep{Hurd03first-orderproof}, and Nitpick \citep{Nitpick}.   Our
formalizations\footnote{The   formalizations are available in the
subdirectories \url{Anderson}, \url{Hajek} and \url{Bjordal} at \url{gi
thub.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/
}.} employ the embedding of higher-order modal logic (HOML) in
classical higher-order logic (HOL) as introduced in previous work
\citep{C40,J30,J23}. We explored the effect of different domain
conditions on the provability of lemmas, theorems and even axioms.
This was motivated by a controversy between Hájek and Anderson
regarding the redundancy of some axioms in Anderson's emendation. In
\emph{constant domain semantics}, the individual domains are the same
in all possible worlds. In \emph{varying domain semantics}, the
domains may vary from world to world. This variation is technically
encoded with the help of an existence relation expressing which
individuals actually exist in each world. Quantifiers are then
uniformly formalized as \emph{actualistic quantifiers} (i.e. guarded by
the existence relation). Our main results are summarized here.

@@

The redundancy of A4 and A5 is particularly controversial. Magari \cite{Magari1988} claimed that A4 and A5 are superfluous\footnote{
  An axiom $A$ is superfluous w.r.t. a set of axioms $S$ iff T3 is entailed by $S \setminus \{ A \}$.
}, arguing that T3 is true in all models of the other axioms and definitions by Gödel. \citet[p.~5-6]{Hajek_Magari_and_others_1996} investigated this further, refuted Magari's claim, but claimed that A4 and A5 are indeed superfluous under the assumption of an additional axiom (PEP, pointwise equality for positiveness), which Magari probably assumed silently. Moreover, \citet[p.~2]{Hajek_Magari_and_others_1996} cites his earlier work\footnote{Although \citep{Hajek_der_Mathematiker_2001} precedes \citep{Hajek_Magari_and_others_1996} in writing, it was published only 5 years later, in German.} \citep{Hajek_der_Mathematiker_2001}, where he claims (in Theorem 5.3) that for Anderson's emended theory \citep{anderson90:_some_emend_of_goedel_ontol_proof}, A4 and A5 are not only superfluous, but also redundant. \citet[footnote 1 in p.~1]{AndersonGettings}, in a footnote, rebutted Hájek's claim, arguing that the redundancy of A4 and A5 holds only under constant domain semantics, while Anderson's emended theory ought to be taken under Cocchiarella's semantics \citep{Cocchiarella} (a varying domain semantics). Our results show that Hájek was originally right, under both constant and varying domain semantics.

Nevertheless, \citet[p.~7]{Hajek2002} (in page 7) acknowledges Anderson's rebuttal, and apparently accepts it, as evidenced by  his use\footnote{
  A4 and A5 are used by \citet[p.~11]{Hajek2002} (page 11) in the proofs of, respectively, Lemma 4 and Theorem 4.
} 
of A4 and A5, as well as varying domain semantics, in his new
emendation (named \AOEH \citep[sec.~4]{Hajek2002}, cf. Appendix
\ref{apx:Hajek1}), which replaces Anderson's A:A1 and A2 by a simpler
axiom H:A12. Surprisingly, the computer-assisted formalization of
\AOEH\ shows that A4 and A5 are still superfluous (not needed for proving T3). Moreover, A4 and A5 are independent\footnote{
  An axiom $A$ is \emph{independent} of a set of axioms $S$ iff there are models of $S$ where $A$ is true and other models of $S$ where $A$ if false. 
} of the other axioms and definitions. Therefore, A4 and A5 are not redundant, despite their superfluousness.

Although, Hájek did not notice the superfluousness of A4 and A5 in his \AOEH, he did describe yet another emendation (his \AOEHH, cf. Appendix \ref{apx:Hajek2}) where A4 and A5 are superfluous (though no claim is made w.r.t. to their redundancy), if A3 is replaced by a stronger axiom (H:A3) additionally stating that the property of actual existence is positive \cite[sec.~5]{Hajek2002}. Formalization of \AOEHH\ shows that A4 is not only superflous, but also redundant. For A5, no conclusive results were achieved; neither a proof nor a countermodel could be automatically generated. Surprisingly, a countermodel for the weaker A3 was succesfully generated. This is somewhat unsatisfactory (for theistic goals), because it shows that \AOEHH\ does not entail the positiveness of being God-like.

Nevertheless, \AOEHH\ is explicitly regarded by \citet[p.~12]{Hajek2002} (page 12) as just an intermediary step towards a more natural theory, based on a more sophisticated notion of positiveness. That is his final emendation (\AOEHHH, cf. \ref{apx:Hajek3}), which restores A3 and does use A4 and A5, albeit in a modified form (i.e. H:A4 and H:A5). The formalization of \AOEHHH\ shows that H:A4 is independent. The old A5 is independent as well, and both H:A4 and H:A5 are superfluous, but no conclusive results were achieved regarding independence or redundancy of H:A5.

@@

\sloppy 
\printbibliography
